Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-07-05 manuscript/manuscript.Rmd
finished the first section.

In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped identifying multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapbale of inferring how decontexualized law-like statements interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acqusition. After all, these experimental results can only be as robust as its experimental settings.

@@

On the other hand, actualistic experiments pay more attention to how experimental results can be generated to archaeological samples. Unlike controlled experiment, variability could be easily observed in actualistic experiment by design, and it cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023]. It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020]. Lastly, actualistic experiment can act as a hurestric for hypothesis generation, echoing the spirit stated by Lin et al. [-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".

@@

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage.

@@

It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics. One such example is the the edge angle [@crabtree1977]

Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].

>>>>

In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped mapping out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition. After all, these experimental results can only be as robust as its experimental settings.

@@

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certian degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment. Unlike controlled experiment, variability could be easily observed in actualistic experiment by design. This feature is crutial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

@@

As implied in its name, the implementation of Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999]. The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].

@@

It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' efforts of reverse engineering during the 19th century, rather than from the intergenerational transmission of knapping knowledge spanning millions of years. This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61]. It is more likely for them to come up with ideas that may not be optimal according to the principles of ergonomics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. Crabtree demonstrated how , challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking. Li
